Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-03-02 VegOzone.tex
Fix hyperref
Restructured Section 2 and spell-ckeck.

\runningtitle{Subarctic biomes for land surface modelling}

@@

Ground level ozone is a highly toxic pollutant known to cause damage to a variety of ecosystems around the World \citep{PT:Emberson2020}. At continuously high concentration levels, ozone negatively affects life expectancy of humans \citep{WHO2008} and is a cause of visible injury, photosynthetic damage, early senescence as well as programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. {\bf TODO: Wealth of evidence demonstrating ozone damage to vegetation (perhaps focus on key evidence from northern climates and key species (e.g. birch, Norway srpuce etc... - this might help to make it clear why we focus on these species with the PODy modelling))}
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation and industrial sectors have increased in the past few decades. Even though long-term (pre 1950) observations of surface ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern Hemisphere have at least doubled and continued to increase until relatively recently \citep[Chapter 2]{IPCC2013}. Recent trend analyses indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level-off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring which saw a move from coal to gas, reduced the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, there are indications that an increasing frequency of thermal stress accompanying climate change will reduce vegetations' capacity for ozone uptake and therefor scotch these achievments \citep{NCC:Lin2020}.\\

@@

In the annual cycle in boreal and temperate climates, a distinct maximum of ground level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. Several conditions unique to the Actic are stimulating the formation of this spring peak. Major pathways of ozone and its precursors' removal from the boundary layer include dry deposition on vegetated surfaces and photochemistry \citep{RG:Clifton2020}. As dry deposition to bare ground or snow and ice covered surfaces was found to be low \citep{ACP:Helmig2007}, both pathways of removal are substantially suppressed during polar night \citep{AE:Monks2000}. This leads to a build up of ozone and its precursors in the course of winter. Come spring precursors are photochemical activated again, while snow cover typically prevails for a longer period of time continuing to suppress ozone dry deposition.
On the other hand, very high ozone concentrations occuring over shorter periodes of time can often be attributed to deep tropopause folding events over the Actic during winter. These events facilitate an intrusion of stratospheric air masses which are enriched in ozone \citep{JGR:Skerlak2015}.
The spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a  high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is the complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as described above as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

The start of growing season in northern Fennoscandia has been shifting to earlier dates in response to the general warming trend \citep[e.g.,]{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007} and thereby converging with the period of the ozone spring peak. At the same time the growing season is also becoming longer. In conjunction with midnight sun conditions (almost $24\,\unit{h}$ photosynthetic activity), a longer growing season, is prolonging the time in which vegetation can accumulate ozone. These factors may promote a higher potential susceptibility of northern vegetation to ozone in the future.

Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange allowing the entry of \chem{CO_2} for photosynthesis and loss of \chem{H_2O} vapour via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the course of the day and growing season primarily to balance \chem{CO_2} uptake against \chem{H_2O} vapour loss. The higher the stomatal conductance, the higher the potential for ozone uptake. Stomatal conductance has been emperically linked to environmental factors such as air temperature, photosynthetic active radiation, vapor pressure deficit, and soil water potential as well as photosynthesis itself.\\

To measure the risk of damage inflicted to vegetation through ozone, different metrics have been defined. Two main types can be differentiated, concentration-based or flux-based metrics, e.g., Accumulated Over a Threshold $\mathrm{x}$ ($\mathrm{AOT_x}$) and Phyto-toxic Ozone Dose over a threshold $\mathrm{y}$ ($\mathrm{POD_y}$), respectively. Due to their simplicity, concentration-based metrics are still widely used. Despite comparisons of exposure-based methods with previously used concentration based methods (e.g. $\mathrm{AOT_{40}}$) show that $\mathrm{POD_y}$ is better suited to represent the risk to vegetation from ozone, especially in the more climate extreme parts of Europe \citep{GCB:Mills2010,ICP:MappingManual2017}.

@@

For computing the $\mathrm{POD_y}$, an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf. Details specific to the $\mathrm{DO_3SE}$ model are given in Section~\ref{sec_do3se}. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged ambient air ozone flux through the stomata Eq.~(\ref{eq:flux_stomata}) and a flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ is representing the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$).

@@

@@

These methods described above rely on accurate representation of plant physiological response to climate variables (i.e. the species or plant funtional type specific responses used to parameterise risk assessment models) and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as sub-arctic climates has not been perfromed and relies on assuming that less extreme boreal parameterizations will be applicable.
It is important to understand the implications of using these more generic parameterizations for more extreme climate regions, here we explore how risk estimates might be over or underestimated using generic vs bespoke sub-arctic parameterisations. we also explore the role [O3] profile and changing climate variables to understand how this risk might alter in the future with the advent of climate change. This allows both an understanding of how the threat from ozone may change over time but also which are the key aspects of plant physiology that might determine risk (and hence which parameters might require special attention to get right) both for pollution impact assessment, but also more generally for describing gas exchange response to changing climates that will influence bio-geochemcial cycling. This might be one key to solve long-standing issues of coupled earthsystem models (ESM) which generally lack PFTs representing sub-arctic conditions in their landsurface model.

Within the scope of our project to assess probable effects of midnight sun conditions on ozone uptake by plants and resultant enhanced susceptibility to ozone inflicted damage, we conducted ambient air ozone concentration measurements at Svanhovd in the growing seasons of 2018/19 in collaboration with the Norwegian Institute for Air Research (NILU). 
In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019. In terms of $\mathrm{AOT_{40}}$, the threshold for protection of vegetation of $3000\,\unit{ppb}$ (integrated over the growing season) given by the EU directive has been well exceeded at many ozone monitoring sites in Southern Norway in the 2018 growing season \citep{NILU2019} but not at Svanhovd.

In this study, we investigate potential main drivers for the observed ozone damage in the growing season 2018 at NIBIO Environment Centre Svanhovd in contrast to 2019. We compare $\mathrm{POD_y}$ values for natural and semi-natural vegetation for derived climatological averages of environmental conditions at Svanhovd as well as for the growing seasons 2018/19 which shall serve as case studies.

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive climatologies for surface ozone concentration at Svanhovd and northern Fennoscandia based on in situ observations and also compute climatologies of surface temperature, precipitation, and global radiation for Svanhovd. Based on these, we evaluate anomalies of environmental key variables in 2018/19 by statistical means. In addition, the derived surface ozone concentration climatology for northern Fennoscandia will be compared with available ozone reanalysis products. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate main environmental drivers for the observed ozone damage in 2018 (Section~\ref{sec:do3se}). In this context, we introduce the$\mathrm{DO_3SE}$ model equations and derive bespoke parameterizations for sub-arctic species to assess systematic uncertainties arising from the use of default parameterizations. In Section~\ref{sec:conc} finally, we will summarize our results, draw conclusions, and give an outlook on further work.

@@

In our assessment of ozone damage on vegetation in the subarctic in relation to the 2018 heatwave, we focus on the NIBIO operated Environment Centre Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley (Section~\ref{subsec:svanhovd}). Throughout the paper, we will use Svanhovd synonymously for the atmospheric monitoring site at Svanvik. The available meteorological monitoring data are described in Section~\ref{subsubsec:atmo_svanvik}. In Section~\ref{subsubsec:ozone_garden}, the ozone garden -- a plant-based ozone monitoring -- will be described in more detail. Other data used for evaluation of the ozone data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use as well reanalysis products (Section~\ref{subsubsec:ozone_rea}) as complement when observational data was not available.

\subsection{NIBIO Environment Centre Svanhovd}
\label{subsec:svanhovd}
"Svanhovd is located beside the Pasvik river, in the middle of the wedge of Norwegian land separating Russia and Finland in the north. It is where the eastern Siberian taiga meets the western Boreal forest. [...] Svanhovd was originally established as a State Demonstration and Experimental Farm in the 1930's, in order to expand agriculture in Pasvik valley [...]" \citep[Svanhovd][]{NIBIO_Svanhovd}.

The locations of the atmospheric monitoring site as well as the ozone garden with respect to the main buildings of the Svanhovd Environment Centre are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}. The Centre is located south of the settlement of Svanvik, where as the border to Russia, the Pasvik river, lies only a few hundred meters to the east. The original farm buildings, protected by law, are located in the lower center, while the Svanhovd Environment Centre follows to the north-east. Left of the Centre and north of the original farm buildings a botanical garden can be found. It is one amongst three such gardens north of the Polar Circle. The Svanhovd Environment Centre focuses on research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

For the purpose of relating observed damage on ozone sensitive plants (e.g. clover and tobacco) to local ozone concentrations, an ozone monitor had been installed during the growing seasons 2018/19. The relevant data are described in more detail in Section~\ref{subsubsec:atmo_svanvik}.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{svanhovd_researchstation}
  \caption{NIBIO Environment Centre Svanhovd close by the settlement of Svanvik, Norway. Atmospheric monitoring site and ozone garden have been marked. Aerial photography \copyright Norges Kartverk.}
  \label{fig:svanhovd_research_station}
\end{figure}

\subsubsection{Atmospheric monitoring}
\label{subsubsec:atmo_svanvik}
For the growing seasons 2018/19, monitoring data of key variables (temperature, pressure, precipitation, and ozone) have been obtained from luftkvalitet.no which is operated by \citet{NILU_AIRQ}.

All agrometeorological variables including temperature and precipitation are available from September 1992 to present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

Relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. The hatched areas mark times when no ozone data was recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, for data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) which occurred from July 12--29 and destroyed about $18000\,\unit{hectare}$ of forest \citep{SOU2019}. However, due to the above mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.

@@

\subsubsection{Ozone garden}
\label{subsubsec:ozone_garden}
An ozone garden consists of carefully selected plant species of which one cultivar is more sensitive to ozone than the other. At the NIBIO Environment Centre Svanhovd, the likely northern-most ozone garden to date, an ozone garden was installed in the course of our project. Cultivated species were, e.g. clover, tobacco, and beans. As shown in Fig.~\ref{fig:ozone_garden}, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

\begin{figure}
  \includegraphics[width=8.3cm]{IMG_7739.JPG}
  \caption{Clover in the Svanhovd ozone garden. The plants had to be secured against herbivores with a wire-mesh fence. The plants shown are approximately $6-12\,\unit{cm}$ in height.}
  \label{fig:ozone_garden}
\end{figure}

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we have to derive a surface ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated by NILU at Svanhovd, but no ozone measurements have been conducted for the past $14\,\unit{years}$. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we also have to have a look at ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are sparse in this region. An overview over the geographical locations of conducted atmospheric monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data is described in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which is not available at all times and spatially sparse, we also look at various ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observations are located on higher ground than Svanhovd which will result in higher surface ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at close vicinity it varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above tree line (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass (c.c. Fig.\ref{fig:svanhovd_research_station}).

>>>>

\runningtitle{Subarctic biomes for land surface modeling}

@@

Ground level ozone is a highly toxic pollutant known to cause damage to a variety of ecosystems around the World \citep{PT:Emberson2020}. At continuously high concentration levels, ozone negatively affects life expectancy of humans \citep{WHO2008} and is a cause of visible injury, photosynthetic damage, early senescence as well as programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone.
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation and industrial sectors have increased in the past few decades. Even though long-term (pre 1950) observations of surface ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase until relatively recently \citep[Chapter 2]{IPCC2013}. Recent trend analyses indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level-off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring which saw a move from coal to gas, reducing the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, there are indications that an increasing frequency of thermal stress accompanying climate change will reduce vegetation's' capacity for ozone uptake and therefor scotch these achievements \citep{NCC:Lin2020}.\\

@@

In the annual cycle in Boreal and temperate climates, a distinct maximum of ground level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. Several conditions unique to the Arctic are stimulating the formation of this spring peak. Major pathways of ozone and its precursors' removal from the boundary layer include dry deposition on vegetated surfaces and photochemistry \citep{RG:Clifton2020}. As dry deposition to bare ground or snow and ice covered surfaces was found to be low \citep{ACP:Helmig2007}, both pathways of removal are substantially suppressed during polar night \citep{AE:Monks2000}. This leads to a build up of ozone and its precursors in the course of winter. Come spring precursors are photochemical activated again, while snow cover typically continues to prevail and suppress ozone dry deposition.
On the other hand, very high ozone concentrations occurring over shorter periods of time can often be attributed to deep tropopause folding events over the Arctic during winter. These events facilitate an intrusion of stratospheric air masses which are enriched in ozone \citep{JGR:Skerlak2015}.
The spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a  high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is the complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as described above as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

The start of growing season (GS) in northern Fennoscandia has been shifting to earlier dates in response to the general warming trend \citep[e.g.,]{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007} and thereby converging with the period of the ozone spring peak. At the same time the growing season is also becoming longer. In conjunction with midnight sun conditions (almost $24\,\unit{h}$ photosynthetic activity), a longer growing season, is prolonging the time in which vegetation can accumulate ozone. For a natural and semi-natural vegetation already subject to rapid climatic changes, these factors could promote a higher potential risk for northern vegetation to ozone induced damage in the not so distant future.

Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange allowing the entry of \chem{CO_2} for photosynthesis and loss of \chem{H_2O} vapor via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the course of the day and growing season primarily to balance \chem{CO_2} uptake against \chem{H_2O} vapor loss. The higher the stomatal conductance, the higher the potential for ozone uptake. Stomatal conductance has been empirically linked to environmental factors such as air temperature $T$, photosynthetic photon flux density (PPFD), vapor pressure deficit (VPD), and soil water potential (SWP) as well as photosynthesis itself. Results from open top chamber (OTC) experiments on downy birch (\emph{Betula pubescens}) and mountain birch (\emph{Betula pubescens toruosa}), native to subarctic regions, indicated reductions in both biomass, in root:shoot ratio, and visible leaf damage under elevated ozone treatment ($\left<\chem{[O_3]}\right> = 36-54\,\unit{ppb}$) \citep{Amb:Manninen2009}. Although considered to be more ozone tolerant due to an absence of visible injuries \citep{Amb:Girgzdiene2009}, \citet{Amb:Manninen2009} also found chlorophyll:carotenoid ratio and polyamines reductions in Scots pine (\emph{Pinus sylvestris}).\\

To measure the risk of damage inflicted to vegetation through ozone, different metrics have been defined. Two main types can be differentiated, concentration-based and exposure-based metrics, e.g., Accumulated Over a Threshold $\mathrm{x}$ ($\mathrm{AOT_x}$) and Phyto-toxic Ozone Dose over a threshold $\mathrm{y}$ ($\mathrm{POD_y}$). Due to their simplicity, concentration-based metrics are still widely used, despite comparisons between the two methods (e.g. $\mathrm{AOT_{40}}$ with $\mathrm{POD_3}$) show that exposure-based metrics are better suited to represent the actual risk to vegetation from ozone, especially in the more climatically extreme parts of Europe \citep{GCB:Mills2010,ICP:MappingManual2017}. Most of these previous studies focused on the Mediterranean region.

@@

For computing the $\mathrm{POD_y}$, an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf. Details specific to the $\mathrm{DO_3SE}$ model are given in Section~\ref{sec:do3se}. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged ambient air ozone flux through the stomata Eq.~(\ref{eq:flux_stomata}) and a flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ is representing the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$).

@@

Within the scope of this study to assess probable effects of conditions unique to northern Fennoscandia on ozone uptake by plants and resultant enhanced susceptibility to ozone inflicted damage, we conducted ambient air ozone concentration measurements at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd in the growing seasons of 2018/19 in collaboration with the Norwegian Institute for Air Research (NILU). 
In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at Svanhovd. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019. In terms of $\mathrm{AOT_{40}}$, the threshold for protection of vegetation of $3000\,\unit{ppb}$ (integrated over the growing season) given by the EU directive has been well exceeded at many ozone monitoring sites in Southern Norway in the 2018 growing season \citep{NILU2019} but not at Svanhovd.

@@

The methods described above rely on accurate representation of plant physiological response to climate variables (i.e. the species or plant functional type specific responses used to parameterize risk assessment models) and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. In this study, conditions at Svanhovd for the growing seasons 2018/19 shall serve as reference.
It is important to understand the implications of using these more generic parameterizations for more extreme climate regions, here we explore how risk estimates might be over or underestimated using generic vs bespoke subarctic parameterization. we also explore the role \chem{[O_3]} variability and changing climate variables to understand how this risk might alter in the future with the advent of climate change. This allows both an understanding of how the threat from ozone may change over time but also which are the key aspects of plant physiology that might determine risk (and hence which parameters might require special attention to get right) both for pollution impact assessment, but also more generally for describing gas exchange response to changing climates that will influence bio-geochemcial cycling. This might be one key to solve long-standing issues of coupled earth system models (ESM) which generally lack PFTs representing subarctic conditions in their land surface model.\\

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive climatologies for surface ozone concentration at Svanhovd and northern Fennoscandia based on in situ observations and also compute climatologies of surface temperature, precipitation, and global radiation for Svanhovd. Based on these, we evaluate anomalies of environmental key variables in 2018/19 by statistical means. In addition, the derived surface ozone concentration climatology for northern Fennoscandia will be compared with available ozone reanalysis products. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate main environmental drivers for the observed ozone damage in 2018 (Section~\ref{sec:do3se}). In this context, we introduce the$\mathrm{DO_3SE}$ model equations and derive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of default parameterizations. In Section~\ref{sec:conc} finally, we will summarize our results, draw conclusions, and give an outlook on further work.

@@

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on the 2018 growing season which was subject to an extended heatwave and its subsequent growing season of 2019. The NIBIO operated Environment Centre Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley serves as main reference. Throughout the paper, we will use Svanhovd synonymously for the atmospheric monitoring site at Svanvik/Pasvik. The available meteorological monitoring data are described in Section~\ref{subsec:atmo_svanvik}. Other data used for cross-evaluation of the ozone monitoring data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use these to evaluate reanalysis products for ground level ozone (Section~\ref{subsubsec:ozone_rea}).

\subsection{NIBIO Environment Centre Svanhovd and atmospheric monitoring}
\label{subsec:atmo_svanvik}
The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The center focuses on research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths. The locations of the atmospheric monitoring site and the ozone garden are marked in an aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). Svanhovd is located south of the settlement of Svanvik, where as the border to Russia, the Pasvik river, lies only a few hundred meters to the east.\\

\begin{figure}[t]
  \includegraphics[width=8.3cm]{svanhovd_researchstation}\\
  \includegraphics[width=8.3cm]{IMG_7739.JPG}  
  \caption{NIBIO Environment Centre Svanhovd close by the settlement of Svanvik, Norway. (a) Atmospheric monitoring site and ozone garden have been marked. Aerial photography \copyright Norges Kartverk; (b) Clover in the Svanhovd ozone garden. The plants had to be secured against herbivores with a wire-mesh fence. The plants shown are approximately $6-12\,\unit{cm}$ in height.}
  \label{fig:svanhovd_research_station}
\end{figure}

An ozone garden was installed in the course of our project at the NIBIO Environment Centre Svanhovd and was likely the northern-most ozone garden to date. An ozone garden consists of selected plant species of which one cultivar is more sensitive to ozone than the other. Cultivated species were, e.g. clover, tobacco, and beans. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

For the purpose of relating observed damage on ozone sensitive plants to local ozone concentrations, an ozone monitor had been installed during the growing seasons 2018/19 in addition to existing monitors of atmospheric key variables (temperature, pressure, precipitation). All variables have been obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. Ozone monitoring data has also been added to EBAS database operated by \citet{NILU_EBAS}.

Long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

All relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data was recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

 As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, for data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) which occurred from July 12--29 and destroyed about $18000\,\unit{hectare}$ of forest \citep{SOU2019}. However, due to the above mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.

@@

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we derive a surface ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated by NILU at Svanhovd, but no ozone measurements have been conducted for the past $14\,\unit{years}$. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we will cross-calibrate our data with long-term ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are sparse in this region. An overview over the geographical locations of conducted atmospheric monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data are described in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which are not available at all times and spatially sparse, we also evaluate various ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long-term observations are located on higher ground than Svanhovd which will result in higher surface ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at close vicinity it varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above tree line (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass (c.c. Fig.\ref{fig:svanhovd_research_station}).
